---
title: "Assignment 3"
author: "Data Science for Biomedical Informatics (BMIN503/EPID600)"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
    
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```   
***

###Instructions
- Download the Rmd version of this file
- Complete the questions below in RStudio using the Rmd file as a template 
- Replace text in brackets with your answers. (There is no need to keep the brackets).
- Save the Rmd document as Assignment3_*YOUR LAST NAME*.Rmd
- Create an HTML version using knitr
- Turn in completed Assignment3_*YOUR LAST NAME*.html file in Canvas under Assignments -> Assignment 3
- Your assignment **must** be in html format or it will not be graded
- Grades will be assigned according to point scheme below and computed as a percentage of total possible points
- Lateness policy: If an urgent and unforseen circumstance comes up and you need an extension to turn in an assignment, please contact Blanca and/or Sherrie as soon as possible. Unless there is an arrangement in place, late assignments will be scored as follows: 
    - 25% of total score if 1 day late
    - 50% of total score if 2 days late
    - 75%  of total score if 3 days late
    - Assignment will not be graded if 4 or more days late
- DUE DATE: 10/5/18


###Final Project - Repository
1. Fork the [Final Project Repository](https://github.com/HimesGroup/BMIN503_Final_Project) to your GitHub account, and then pull this repository as a new project on your local computer by using the _Project_ menu in the upper-right-hand corner of RStudio, selecting a _New Project_, creating it via _Version Control_ -> _Git_ -> and entering your GitHub repository (remember to append `.git` to the copied `url`). You can now work on your final project from your computer, and easily push updates to your GitHub account, where the final version will be saved. A _Git_ tab will be available that you can use to _Pull_, _Commit_, and _Push_ changes between your local version and that on GitHub without having to use the command line. *(1 point for forking repository)*

https://github.com/DataAndDeadlifts/BMIN503_Final_Project

2. In 2-3 sentences, describe what question you will answer for your final project and what data will you use. There is a list of public data sources at the bottom of the [Optional Reading](https://github.com/HimesGroup/BMIN503/blob/master/Module1_Optional_Reading.Rmd) document that you can look through if you are having difficulty identifying public data. If you'd like help to pick a question, please let Blanca or Sherrie know soon. *(2 points)*

> I plan to group and map protein structure characteristics to disease and pathologys. Perhaps predict proteins with known structure but no known link to disease using the classification/grouping model I build.

3. What three faculty/staff (name, title, division/department) have you met or are planning to meet to discuss your project? *(2 points)*

> Dr. Jason Moore (jhmoore@upenn.edu), Professor and Director of Informatics in the Department of Biostatistics and Epidemiology
> Roland L. Dunbrack (Roland.Dunbrack@fccc.edu), Jr. Ph.D., Adjunct Professor of Biochemistry & Biophysics
> Dr. Graciela Gonzalez (gragon@upenn.edu), Professor of Computer and Information Science 

### Cleaning and Transforming data, Descriptive Statistics and Regression
4. Install and load the `nycflights13` package to do this problem. This package includes the [flights](https://www.rdocumentation.org/packages/nycflights13/versions/1.0.0/topics/flights) dataset, which contains information on 336,776 flights that departed from New York City in 2013 and is often used to learn `dplyr` commands. If you need more guided practice with `dplyr`, beyond what we did in class, you can go through this [vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html). Use `dplyr` to answer the following questions by inserting code below each bullet to provide the answer directly. *(10 points)*

```{r}
#install.packages("nycflights13")
#install.packages("dplyr")
#install.packages("ggplot2")
library("dplyr")
library("ggplot2")
```


    + How many carriers departed NYC in 2013? Which carrier had the most departing flights?
  
```{r}
# NYC airports include JFK, EWR and LGA
flights <- nycflights13::flights
# Num carriers departed NYC in 2013
n_carriers = nrow(distinct(flights['carrier']))
print("Num carriers that departed NYC in 2013;")
print(n_carriers)

# Are there any origins beside NYC?
distinct(flights['origin'])
# Looks like there isn't, no need to filter then!
carrier_flight_count <- table(flights$carrier)
max_carrier <- names(which(carrier_flight_count==max(carrier_flight_count)))
print("Carrier with most flights")
print(max_carrier)
```

    
    + Of the flights that departed in June and July, how many unique tailnum contained the letter M or P?
    
```{r}
# Get unique tailnums in June, July flights
unique_tails <- distinct(filter(flights, month==6 | month==7)['tailnum'])
# Get tails that contain M or P
m_or_p_tails <- filter(unique_tails, grepl("M",tailnum) | grepl("P",tailnum))
print("Number of tailnums from flights in june and july that contain the letter M or P.")
print(nrow(m_or_p_tails))
```
  

    + What was the most popular destination each month?
    
```{r}
pop_dest_by_month <- summarize(group_by(flights,month,dest),count =n()) %>%
  filter(count==max(count))
pop_dest_by_month
```
    
    + Was the mean departure delay time per carrier related to the total number of flights per carrier? The answer should show a plot and use one sentence to address this question qualitatively.
    
```{r}
count_carrier_delay <- summarize(group_by(flights,carrier),count =n(), mean_dep_delay=mean(dep_delay, na.rm=T))

ggplot(count_carrier_delay, aes(x=count, y=mean_dep_delay)) +
  geom_point(size=2) + 
  geom_smooth(method="loess",se=TRUE,fullrange=FALSE, level=0.95) +
  geom_text(label=count_carrier_delay$carrier, nudge_x=1500)
```

There doesn't appear to be any kind of clear link between number of flights and mean departure delay based on this data. Reasonable hypotheses before seeing this data may be either 1. Mean Dept delay increases with flight count - as there are more opportunities for errors to be made in organizing the flight... or 2. Mean Dept delay may decrease with flight count, as the number of flights increases the expertise of the airline, leading to better organization of flights. Neither of these hypotheses are supported by the data for NYC in 2013. There is no strong and clear linear relationship between flight count and mean departure delay.
    
    + Was the mean delay time for a given time of departure related to the time of departure? The answer should show a plot and use one sentence to address this question qualitatively.
    
```{r}
time_of_day_delay <- summarize(group_by(flights,sched_dep_time), mean_dep_delay=mean(dep_delay, na.rm=T))

ggplot(time_of_day_delay, aes(x=sched_dep_time, y=mean_dep_delay)) +
  geom_point(size=2) + 
  geom_smooth(method="auto",se=TRUE,fullrange=FALSE, level=0.95)
```

The method of regression against these variables displays a high level of confidence in it's fit. This belies a strong relationship between sched_dep_time and mean_dep_delay. Around 1800-1900 during the day there is a much higher average delay time paired with a high confidence (low range) confidence interval over that period. If I were flying out of NYC in 2013 I would definitely avoid 7PM and instead try to schedule flights that depart at 5AM if I really don't want to be delayed.


5. Load the `MASS` package, which contains the  [birthwt](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/birthwt.html) dataset, to answer the following questions. We will work toward identifying variables that are associated with low birthweight, both as a dichotomous outcome and a continuous one. HINT: The `MASS` package contains a `select` command that may compete with that from `dplyr`. If you have `dplyr` and `MASS` loaded simultaneously and want to use `dplyr`'s `select`, do it explicitly. For example, use `dplyr::select` rather than `select`. *(9 points)*

```{r}
#install.packages("MASS")
library("MASS")
```


    + Take a look at the `birthwt` dataset (e.g., using `str()`) to get a sense of variables it contains. We will use the variables `low` and `bwt` as outcomes and `age`, `race`, `smoke`, `ht`, `ptl` as predictors. Create a new dataset `birthwt.clean` containing only the variables we are interested in, where;
    (1) the variables `low`, `ht`, and `ptl` are renamed `low.birth.wt`, `hypertension`, and `prem.labor`, respectively, 
    (2) all integer categories (i.e., `low.birth.wt`, `race`, `smoke`, `hypertension`, `prem.labor`) are converted to factors with intuitive labels, 
    (3) the `race` variable does not include the `other` category, and 
    (4) the `ptl` variable is dichotomized into two levels based on a positive or negative history of previous premature labor. 
    
```{r}
birthwt <- birthwt
#str(birthwt)
# 1
birthwt.clean <-  dplyr::select(birthwt,low.birth.wt=low, birth.wt=bwt, age, race, smoke, hypertension=ht, prem.labor=ptl)
# 2
integer_cols <- c('low.birth.wt','race','smoke','hypertension','prem.labor')
birthwt.clean$low.birth.wt <- factor(birthwt.clean$low.birth.wt, labels=c("normal","low"))
birthwt.clean$race <- factor(birthwt.clean$race, labels=c("white","black","other"))
birthwt.clean$smoke <- factor(birthwt.clean$smoke, labels=c("non-smoking","smoking"))
birthwt.clean$hypertension <- factor(birthwt.clean$hypertension, labels=c("no-history","history"))
# 3
birthwt.clean <- filter(birthwt.clean, race!="other")
# 4
birthwt.clean$prem.labor <- factor(ifelse(as.numeric(as.character(birthwt.clean$prem.labor))==0, 0, 1))
birthwt.clean$prem.labor <- factor(birthwt.clean$prem.labor, labels=c("no-history","history"))

str(birthwt.clean)
```
    
    + Perform univariate analysis for the resulting `age` and `smoke` variables you created as predictors and `low.birth.wt` and `bwt` as outcomes. Include appropriate plots that look at the pairwise relationships between the outcomes and each predictor. Are there differences between the results for each outcome?
    
```{r}
# Smoke vs low.birth.wt and birth.wt
ggplot(birthwt.clean, aes(smoke)) +
  geom_bar(aes(fill=low.birth.wt))

ggplot(birthwt.clean, aes(x=smoke, y=birth.wt)) +
  geom_boxplot()

# Age vs low.birth.wt and birth.wt
ggplot(birthwt.clean, aes(x=low.birth.wt, y=age)) +
  geom_boxplot()

ggplot(birthwt.clean, aes(x=age,y=birth.wt)) +
  geom_point() +
  geom_smooth(method="auto",se=TRUE,fullrange=FALSE, level=0.95)
```
    
```{r}
# Smoke x low.birth.wt contingency table
smoke_by_lowbirthwt <- table(birthwt.clean[, c('smoke', 'low.birth.wt')])
print("Smoke X low.birth.wt contingency table test (chi-squared):")
chisq.test(smoke_by_lowbirthwt)
# Smoke x birth.wt t-test
x1 = filter(birthwt.clean, smoke=='non-smoking')$birth.wt
x2 = filter(birthwt.clean, smoke=='smoking')$birth.wt
print("Smoke x birth.wt t-test:")
t.test(x1, x2)
# low.birth.wt x age t-test
n1 = filter(birthwt.clean, low.birth.wt=='normal')$age
n2 = filter(birthwt.clean, low.birth.wt=='low')$age
print("Age x low.birth.wt t-test:")
t.test(n1, n2)
# birth.wt x age spearman correlation
print("Age x birth.wt spearman correlation test")
cor.test(birthwt.clean$age, birthwt.clean$birth.wt, method="spearman")
```

(1) Smoking and low.birth.wt. It appears that smoking and low birth weight have a statistically significant association with a p-value of 0.003537. Smoking likely has an effect on whether a baby is born with a low birth weight.

(2) Smoking and birth.wt. It appears according to the data that smoking affects birth weight at a statistically significant level (0.05 p-value cutoff). The distribution of birth weights have a significantly different mean when split by smoking status of the mother. Smoking is therefore likely to have a harmful effect on birth weight according to this data.

(3) Age and low.birth.wt. According to this data age does not have a statistically significant effect on birth weight at a 0.05 significance level. Age likely does not have a significant effect on low birth weight.

(4) Age and birth.wt. According to this data age does not have a statistically significant effect on birth weight at a p-value cutoff of 0.05. Age is not likely to have an effect on birth weight according to this data.
    
    + Create multivariable regression models for `low.birth.wt` and `bwt` using `age`, `race`, `smoke`, `hypertension`, and `prem.labor` as predictors. What factors are significantly associated with the birth weight outcomes?

```{r}
lowbwt_model <- glm(low.birth.wt ~ age + race + smoke + hypertension + prem.labor, birthwt.clean, family='binomial')

print("Low birth weight model:")
summary(lowbwt_model)

bwt_model <- lm(birth.wt ~ age + race + smoke + hypertension + prem.labor, birthwt.clean)

print("Birth weight model:")
summary(bwt_model)
```

If I'm interpreting this correctly, age negatively impacts the babies likelihood in having a low birth weight and all other factors (race being black, smoking, hypertension and premature labor history) all increasing the likelihood of a low birth weight. Of these factors, a history of premature labor in the mother is the highest estimator in a low birth weight.

The birth weight model esimates that all factors except age predict a lower birth weight with their being "positive" or present (as if happening, ex the mother smoking or the mom having hypertension). Given these factors, smoking seems to have the highest predictive and negative impact on the resulting birth weight.
