---
title: "Assignment 5"
author: "Data Science for Biomedical Informatics (BMIN503/EPID600)"
output:
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---

```{r global options, include=FALSE}
library(tidyverse)
library(sf)
library(tidycensus)
library(leaflet)

knitr::opts_chunk$set(warning=FALSE, message=FALSE)
``` 

***

####Instructions
- Download the Rmd version of this file
- Complete the questions below in RStudio using the Rmd file as a template 
- Replace text in brackets with your answers. (There is no need to keep the brackets).
- Save the Rmd document as Assignment5_*YOUR LAST NAME*.Rmd
- Create an HTML version using knitr
- Turn in completed Assignment5_*YOUR LAST NAME*.html file in Canvas under Assignments -> Assignment 5
- Your assignment **must** be in html format or it will not be graded
- Grades will be assigned according to point scheme below and computed as a percentage of total possible points
- Lateness policy: If an urgent and unforseen circumstance comes up and you need an extension to turn in an assignment, please contact Blanca and/or Sherrie as soon as possible. Unless there is an arrangement in place, late assignments will be scored as follows: 
    - 25% of total score if 1 day late
    - 50% of total score if 2 days late
    - 75%  of total score if 3 days late
    - Assignment will not be graded if 4 or more days late
- DUE DATE: 11/8/18


#### Final Project - Overview and Introduction
1. Recall that you forked the [Final Project Repo](https://github.com/HimesGroup/BMIN503_Final_Project) and have downloaded it as a project to your local computer. Write the overview and introduction for your final project. The overview consists of 2-3 sentences summarizing the project and goals. For the introduction, the first paragraph describes the problem addressed, its significance, and some background to motivate the problem. In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

> Replace this text with the overview and introduction but keep in mind that it should be part of your GitHub Repo for the final project so your working version should be stored there.

#### Static Maps
2. Create maps of county-level obesity rate estimates for adults living in the contiguous United States using BRFSS data from 2004 and 2014. These estimates have already been age-adjusted using Census population estimates to allow for comparison between counties and across time.
    + Read in [BRFSS obesity data](https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/county_obesity_prevalence.csv) and [county polygons](https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/uscounties_2010.rds), and name them `obesity` and `counties`, respectively. Use the base _plot_ function to check that `counties` includes the polygon elements you expect. Hint: reading in an RDS file from a website requires that you run the file through a _decompressor_ before loading it via `readRDS`. R has a built-in decomopressor function called `gzcon`. *(2 points)*
    + What were the 2004 and 2014 obesity rates for Orange County, California? For Orange County, Texas? Show all variables associated with these counties in the BRFSS and county polygons datasets. Aside from county names, what identifiers do these datasets share? *(2 points)*
    + Merge the two datasets so that `counties` contains state names and obesity rates for 2004 and 2014. *(3 points)*
    + For each year (i.e., 2004 and 2014), create a static choropleth map of county-level obesity rates for the US using _ggplot2_. Add a title with `ggtitle`, remove county borders with `lwd=0` in the `geom_sf` call, and incorporate custom theme elements with the user-created `my_theme()` function. Some code to get you started with these maps is offered below. Feel free to change plot aesthetics or choose a different color palette. Hint: be sure to set `eval = TRUE` for your map to load! *(4 points)*
    + How did obesity rates in adults change between 2004 and 2014? (Qualitative answer is sufficient!) *(2 points)*    
    
```{r eval = FALSE}
library(RColorBrewer)
library(leaflet)

# Use a fixed color scale to more easily compare obesity rates between maps 
prev_min <- YOUR CODE TO FIND MINIMUM OBESITY RATE
prev_max <- YOUR CODE TO FIND MAXIMUM OBESITY RATE

my_theme <- function() {
  theme_minimal() +                                  
  theme(axis.line = element_blank(),                 
        axis.text = element_blank(),                 
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  
        legend.key.size = unit(0.8, "cm"),          
        legend.text = element_text(size = 16),       
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 22))      
}

myPalette <- colorRampPalette(brewer.pal(9, "YlOrRd"))

# Replace "..." with your code
ggplot() +
  geom_sf(..., lwd = 0) +
  ... +
  scale_fill_gradientn(name = "Obesity rate (%)", colours = myPalette(100),
                       limit = range(prev_min, prev_max)) 

```

#### Interactive Maps
3. Create an interactive map to visualize the change in adult obesity rates for all counties in the contiguous United States between 2004 and 2014.
    + Create a new variable in `counties` that tracks the _change_ in obesity rate for each county between 2004 and 2014. Be sure to code this variable so that a positive value indicates an increase in the prevalence of obesity. *(1 point)*
    + Create an interactive choropleth map using the _leaflet_ library to visualize changes in county-level obesity rates between 2004 and 2014. Be sure to include a legend and scalebar for your map. It may be helpful to use the _leaflet_ code from the practicum as a starting point, but be sure to incorporate the palette function provided below to color your map. The popup message for your map should be formatted like the following: *(5 points)*
    
> Philadelphia County, Pennsylvania <br/>
> Change in obesity rate (2004-2014): 2.8% 


```{r eval = FALSE}
library(leaflet)

# Bins continuous variables into an ordinal color scale appropriate for our data
pal_fun <- colorBin(palette = brewer.pal(9, "RdBu")[c(1:5, 7)], 
                    bins = c(-3, -1, 1, 5, 9, 13, 17), reverse = TRUE,
                    NULL)

```

4. Create a choropleth map of a county-aggregated variable of your choice from the American Community Survey (ACS) 5-year estimates for 2012-2016. 
    + Write a line of code to show how you could use a _tidycensus_ command to view variables from the ACS 5-year estimates for 2012-2016. Store the results as an object named `vars`. You do not need to show a preview of `vars`. *(1 point)*
    + Use `get_acs` from _tidycensus_ to query a variable of your choice from the ACS 5-year estimates for 2012-2016 at the county level. For percentages, be sure to obtain both an estimate and a total. For full points, assign the ACS data you obtained to a data frame named `acs.data.` Make sure each county is represented by no more than a single row of `acs.data` and each column of `acs.data` contains no more than a single variable and has a meaningful name. We are only interested in mapping the estimate, so you can remove information about margin of error. Show the first six rows of `acs.data`. It is up to you which variable you would like to map. Possible options include:
        + Poverty rate (estimate: B17010_002, total: B17010_001)
        + Median household income (B19013_001)
        + Median house value (B25077_001) *(4 points)* 
        
    + Merge `counties` with `acs.data` so that `counties` contains the ACS variables you have chosen to map. Show the first six rows of `counties`. *(2 points)*
    + Make an interactive map with _leaflet_. Be sure to include a popup message, legend, and scalebar. An example palette function has been provided for you, but feel free to choose a different color palette. For the popup message, be sure to label "$" or "%" as appropriate *(4 points)*
    + Describe in 1-2 sentences the geographic distribution of your chosen variable across the United States. Where is this rate/value the highest? *(1 point)*
    
```{r}

pal_fun <- colorNumeric("BuPu", NULL)       

```
    
    